A Study of the Effectiveness of Usage Examples in REST API Documentation

### 2017-04-20 controlled_study.tex
adds the suggested text changes by Martin sans the agreement portion

To overcome the shortcomings found from the first pilot study, we decided to develop a web-based REST API explorer as shown on Fig. \ref{fig:api_explorer} that allows participants to use their browser to make the API calls without setting up any development environment. The web-based API explorer only requires the inputs I1-4, and displays the HTTP response headers and body on the click of a button. Thus the participants could focus on using the right input and verifying the output without having to write any code. The web-base REST API explorer also allowed us to automatically collect all the trial API calls that the participants make for each API task. A second pilot study involving seven new participants was performed to collect usability related feedback about the web-based REST API explorer. We only used WordPress since it didn't require knowledge about OAuth. From this study, we observed patterns of mistakes that API client developers make that can be reduced by adding usage examples in the API documentation.

@@

\subsection{Quantitative Analysis Results} % (fold)

We present a summary of the quantitative results as found by analyzing a total of 539 API calls (385 from G1, 152 from G2) made by the participants from this study in Fig. \ref{fig:comparison}. Fig. \ref{graph:productivity} juxtaposes the average number of trial API calls made by each group against their rate of success in performing the given tasks T2-T6. The average number of trial API calls and success rate for each group is computed using the following formulas:

Even though it is expected that G2 would outperform G1 with an enhanced documentation, we were surprised to see the difference in effort vs. success of the two groups. In Fig. \ref{graph:productivity} and \ref{graph:time} we see that, for the simpler tasks, T2 and T3, G1 required more time and more trial API calls and reached successful answers with a lower percentage compared to G2. For T4, where the participants are required to inspect an HTTP response header, G2 needed more API calls on average, but also yielded 24\% higher success rate. For the more complex tasks T5 and T6, we see that participants in G1 had very low success rate and very high number of average API trials compared to G2. The data here depicts that without usage examples developers spend more time and require more trial and error attempts yet resulting in lower success rates, and the situation gets worse for relatively complicated tasks.

@@

Based on these results, we recommend REST API developers to study the usability of their API documentation with prospective API client developers and provide required usage examples in the documentation to improve API client developer productivity and satisfaction.

@@

11 of the 16 G1 participants mentioned the lack of examples with HTTP headers, request and response as the primary problem with the API documentation. On the other hand, 7 out of the 10 G2 participants mentioned the examples being helpful from the enhanced documentation. We identify usage examples as the most frequent and central topic of interest from our participants as per their feedback. This is also supported by our quantitative analysis results as discussed earlier in this section. REST API developers need to provide usage examples showing sample requests and responses with HTTP headers and body for API actions to help API usability. In the remainder of this section, we provide our observations related to each task and their implications.

\textbf{Observation 1. Use of Data Type.} Without usage examples, we observed 10 of the 16 G1 participants used incorrect data type of \lstinline{String} instead of \lstinline{Integer}. To successfully complete T1, participants had to use an integer ID representing an \lstinline{Author} object given the author name. In the original API documentation, the following is mentioned about the author parameter: ``author: Limit result set to posts assigned to specific authors.'' The documentation doesn't mention that the data type required is the numeric author ID, not the author name. P1.12 mentioned the following: \textit{``...to find posts for the author, I had to inspect the response to see that indeed the $author\_id$ was in there''}. The enhanced documentation showed one example of using numeric ID. 3 of the 10 participants from G2 ran into the same error. When faced with this error, participants used trial and error to eventually complete the task.

@@

\textbf{Observation 2. Use of Data Format.} We observed participants in G1 faced obstacles using correct data formats for Date and Array type data in API requests without usage examples. Tasks T3 required the use of a correct format to represent an array of numeric post IDs to be excluded using the API. The original documentation mentions the following about this API query parameter: ``exclude: Ensure result set excludes specific IDs.'' Participants attempted to solve this task using multiple different formats for specifying the array. For example: they tried with \lstinline{exclude=[1,4], exclude=1&4, exclude=4&&exclude=1, id!=1&id!=4} and other alternatives before eventually finding the right format as follows \lstinline{exclude=1,4}. Participant P1.6 mentioned the following: \textit{``It was difficult to figure out whether some of the inputs needed to be arrays or just a comma-separated list.''}. 7 out of the 16 participants in G1 faced this error. The enhanced documentation showed one example of using multiple IDs and all G2 participants could use the right format.

Task T5 also required the participants to use an ISO8601 formatted date for publishing a post with a specific date. Even though the original documentation mentions ISO8601 format, it doesn't provide an example. Instead of using ``2016-01-01T00:00:00-00:00'', participants from G1 used date formats as follows: \lstinline{``January 1, 2016'', ``2016-01-01T12:00:00'', ``20160101''}. P1.8 mentioned the following: \textit{``Got hung up trying to figure out the formatting of the date (which doesn't appear to be ISO8601, despite what it says)''}. After analyzing the response, we found that P1.8 used the ISO8601 formatted date in the API calls, but didn't provide the required \lstinline{second} portion of the time. 8 of the 16 participants in G1 had an error in the date formats. With an example, 1 of the 10 participants had the same error from G2.

@@

\textbf{Observation 3. Use of Request Headers.} We were surprised to see that 14 of the 15 G1 participants that tried T5 and T6 faced problem using correct request headers without usage examples. Tasks T5 and T6 required the use of a HTTP request header named \lstinline{Content-Type}. In the original API documentation, the following was mentioned: ``The API uses JSON exclusively as the request and response format, including error responses.''. Experienced software engineers still had problem using this HTTP header as we found from the participants in G1. P1.12 mentioned the following: \textit{``I didn't know how to specify application/json so that I could send POST data via the body instead in the URL parameters... Examples would really help - with a sample request and a sample response.''}  Even though G1 participants have years of experience with REST APIs, only one of them could use the \lstinline{Content-Type} header correctly without example. In contrast, 2 out of the 10 participants from G2 failed to use this header correctly as indicated in the example in the enhanced documentation. Moreover, 1 of the 2 G2 participants that didn't use the header first time, later used it correctly on a subsequent attempt.

@@

\section{Threats to Validity}
\label{sec:threat}
  The selection of the API, the tasks, and the participants may introduce selection bias. We have developed and followed a set of requirements as discussed earlier on this paper to reduce this bias (see Section \ref{sub:protocol}). This study needs to be replicated with different sets of APIs, tasks, and participants to generalize the results. The analysis of the raw data may introduce a bias. To reduce this, we have involved multiple researchers in the analysis process.

>>>>

To overcome the shortcomings found from the first pilot study, we decided to develop a web-based REST API explorer as shown on Fig. \ref{fig:api_explorer} that allows participants to use their browser to make the API calls without setting up any development environment. The web-based API explorer only requires the inputs I1-4, and displays the HTTP response headers and body on the click of a button. Thus the participants could focus on using the right input and verifying the output without having to write any code. The web-base REST API explorer also allowed us to automatically collect all the trial API calls that the participants make for each API task. A second pilot study involving seven new participants was performed to collect usability related feedback about the web-based REST API explorer. We only used WordPress since it did not require knowledge about OAuth. From this study, we observed patterns of mistakes that API client developers make that can be reduced by adding usage examples in the API documentation.

@@

\subsection{Quantitative Analysis} % (fold)

We present a summary of the quantitative results as found by analyzing a total of 539 API calls (385 from G1, 152 from G2) made by the participants from this study in Fig. \ref{fig:comparison}. Fig. \ref{graph:productivity} juxtaposes the average number of trial API calls made by each group against their rate of success in performing the given tasks T2-T6. The average number of trial API calls and success rate for each group is computed using the aforementioned formulas.

Even though it is expected that G2 would outperform G1 with an enhanced documentation, we were surprised to see the extent of the difference in effort vs. success of the two groups. In Fig. \ref{graph:productivity} and \ref{graph:time} we see that, for the simpler tasks, T2 and T3, G1 required more time and more trial API calls and reached successful answers with a lower percentage compared to G2. For T4, where the participants are required to inspect an HTTP response header, G2 executed more API calls on average, but also yielded 24\% higher success rate. For the more complex tasks T5 and T6, we see that participants in G1 had very low success rate and very high number of average API trials compared to G2. The data here confirms the intuition that without usage examples developers spend more time and execute more trial and error attempts yet have lower success rates. And the situation gets worse for relatively complicated tasks.

@@

The results provide an initial quantification of the impact of examples for typical REST-API usage tasks. This quantification is preliminary evidence that can help justify future investment in API documentation.

@@

In their free-form feedback, 11 of the 16 G1 participants mentioned the lack of examples with HTTP headers, request and response as the primary problem with the API documentation. On the other hand, 7 out of the 10 G2 participants mentioned the examples being helpful from the enhanced documentation. We identify usage examples as the most frequent and central topic of interest from our participants as per their feedback. This is also supported by our quantitative analysis results. In the remainder of this section, we provide our observations related to each task and their implications.

\textbf{Observation 1. Use of Data Type.} Without usage examples, we observed 10 of the 16 G1 participants used incorrect data type of \lstinline{String} instead of \lstinline{Integer}. To successfully complete T1, participants had to use an integer ID representing an \lstinline{Author} object given the author name. In the original API documentation, the following is mentioned about the author parameter: ``author: Limit result set to posts assigned to specific authors.'' The documentation does not mention that the data type required is the numeric author ID, not the author name. P1.12 mentioned the following: \textit{``...to find posts for the author, I had to inspect the response to see that indeed the $author\_id$ was in there''}. The enhanced documentation showed one example of using numeric ID. 3 of the 10 participants from G2 ran into the same error. When faced with this error, participants used trial and error to eventually complete the task.

@@

\textbf{Observation 2. Use of Data Format.} We observed participants in G1 faced obstacles using correct data formats for \lstinline{Date} and \lstinline{Array} type data in API requests without usage examples. Tasks T3 required the use of a correct format to represent an array of numeric post IDs to be excluded using the API. The original documentation mentions the following about this API query parameter: ``exclude: Ensure result set excludes specific IDs.'' Participants attempted to solve this task using multiple different formats for specifying the array. For example: they tried with \lstinline{exclude=[1,4], exclude=1&4, exclude=4&&exclude=1, id!=1&id!=4} and other alternatives before eventually finding the right format as follows \lstinline{exclude=1,4}. Participant P1.6 mentioned the following: \textit{``It was difficult to figure out whether some of the inputs needed to be arrays or just a comma-separated list.''}. Seven of the 16 participants in G1 faced this error. The enhanced documentation showed one example of using multiple IDs and all G2 participants could use the right format.

Task T5 also required the participants to use an the ISO8601 formatted date for publishing a post with a specific date. Even though the original documentation mentions ISO8601 format, it does not provide an example. Instead of using \lstinline{2016-01-01T00:00:00-00:00}, participants from G1 used date formats as follows: \lstinline{January 1, 2016, 2016-01-01T12:00:00, 20160101}. P1.8 mentioned the following: \textit{``Got hung up trying to figure out the formatting of the date (which doesn't appear to be ISO8601, despite what it says)''}. After analyzing the response, we found that P1.8 used the ISO8601 formatted date in the API calls, but did not provide the required \lstinline{second} portion of the time. Eight the 16 participants in G1 had an error in the date formats. With an example, 1 of the 10 participants had the same error from G2.

@@

\textbf{Observation 3. Use of Request Headers.} We were surprised to see that 14 of the 15 G1 participants that tried T5 and T6 faced problem using correct request headers without usage examples. Tasks T5 and T6 required the use of a HTTP request header named \lstinline{Content-Type}. In the original API documentation, the following was mentioned: ``The API uses JSON exclusively as the request and response format, including error responses.''. Experienced software engineers still had problem using this HTTP header as we found from the participants in G1. P1.12 mentioned the following: \textit{``I didn't know how to specify application/json so that I could send POST data via the body instead in the URL parameters... Examples would really help - with a sample request and a sample response.''}  Even though G1 participants have years of experience with REST APIs, only one of them could use the \lstinline{Content-Type} header correctly without example. In contrast, 2 out of the 10 participants from G2 failed to use this header correctly as indicated in the example in the enhanced documentation. Moreover, 1 of the 2 G2 participants that did not use the header the first time, later used it correctly on a subsequent attempt.

@@

\section{Threats to Validity}
\label{sec:threat}
  The selection of the API, the tasks, and the participants may introduce selection bias
 that there are obstacles we don't know about that could be more severe, but were not observed because the tasks did not involve anything related to these obstacles.
  The analysis of the raw data may introduce a bias. To reduce this, we have involved multiple researchers in the analysis process.
